---
title: "Edges, abundances and sun"
author: "Filipey and Piew"
date: "2025-06-17"
output: 
  html_document:
    theme: darkly
    highlight: haddock
   # code_folding: hide  # hide code by default
    toc: true # table of content true
    toc_depth: 4  # upto three depths of headings (specified by #, ## and ###)
    number_sections: false  ## if you want number sections at each table header
    toc_float: true
    css: css_perso.css
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Plan of action
## Idea on what we could do

- **Basic visualisation** 
  - simple representation of the variables relation
  - sample size and distribution of the predictor
  
  <br>
  
- **Univariate model selection** 
  - $y_d$ (decreasing) - `[0, 1]`
  - $y_i$ (increasing) - `[0, 1]`

<br>

- **Multivariate model** 
  - Fixed and random structured selected during univariate model selection
  - Test for bivariate model $(y_d, y_i)$ using `brms`
  - Test for trivariate model with the constant dynamic $y_c$ : $(y_d, y_c, y_i)$

<br>
<br>

## Some considerations

The better would be to go *crescendo* in term of complexity and maybe focus on the modelisation of the demographic dynamics **at the taxa levels** for a first try, and then analyse the specific-slopes. With that in mind, we have to think on the way we'll sort the dataframe. At the **population level**, could be useful to keep all the statistical power we have, but at the deeper (specific) level, maybe filter by species' population sample size.


<br>
<br>

## Data importation
`data` have been pre-treated within the script <script>. 

```{r}
summary(cars)
```
